Alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in  hospitalized COVID-19 patients

### 2022-03-25 manuscript/main_manuscript_revision.Rmd
Pipeline update
Clustering tools provided now by a development package

Severe 
<span custom-style = 'revision'>coronavirus disease 19 (COVID-19)</span> 
manifests with systemic immediate pro-inflammatory innate immune activation and altered iron turnover. As iron homeostasis, 
<span custom-style = 'revision'>differentiation and function of myeloid leukocytes</span> 
are interconnected we sought to characterize the cellularity, surface marker expression and iron transporter phenotype of neutrophils and monocyte subsets in COVID-19 patients within 72 hours from hospital admission, and how 
<span custom-style = 'revision'>these parameters</span> 
relate to infection severity and 
<span custom-style = 'inflammatory response.'>these parameters</span>

Between March and November 2020, blood leukocyte samples from 
<span custom-style = 'revision'>hospitalized</span> 
COVID-19 patients 
<span custom-style = 'revision'>(n = 48)</span> 
and healthy individuals 
<span custom-style = 'revision'>(n = 7)</span> 
were analyzed by flow cytometry enabling comparative analysis of 40 features. 
<span custom-style = 'revision'>Inflammation driven neutrophil expansion, depletion of CD16^+^ non-classical monocytes and changes in surface expression of neutrophil and monocyte CD64 and CD86 were associated with COVID-19 severity. By unsupervised self-organizing map clustering, four patterns of innate myeloid response were identified and linked to varying levels of systemic inflammation, iron witholding and the disease course.</span> These alterations of the myeloid leukocyte compartment 
<span custom-style = 'revision'>during acute COVID-19</span> 
may be hallmarks of inefficient viral control, 
<span custom-style = 'revision'>immune hyper-activation and immunosupression</span> and help at risk prediction and treatment 
<span custom-style = 'revision'>optimization.</span>

@@

Since late 2019, the 
<span custom-style = 'revision'>SARS-CoV-2 pandemic</span> 
has resulted in a global health crisis. 
<span custom-style = 'revision'>As of February 2022, more than 420 million infections and over 5.8 million COVID-19 deaths</span> 
were documented [@WHO2021]. 
COVID-19 is characterized by a heterogeneous clinical manifestation and divergent courses of the disease. This ranges from asymptomatic 
<span custom-style = 'revision'>infection, to ambulatory disease with wide-ranged flu-like and multi-organ complaints, up to severe respiratory or multi-organ condition </span>, 
requiring hospitalization and treatment at intensive care unit [@Sahanic2021; @Oberfeld2020]. 
Especially in the latter cases, an excessive over-stimulation of the immune system, likewise driven by an impaired immune control of viral replication leads to a ‘cytokine release syndrome’ with subsequent organ damage [@Riva2020; @Fajgenbaum2020]. Specifically, low levels of CD4^+^ and CD8^+^ cells, 
<span custom-style = 'revision'>impaired anti-viral type I interferon response</span> 
and polymorphisms in the genes involved in antigen presentation 
<span custom-style = 'revision'>were associated with the risk of severe COVID-19</span> [@Ellinghaus2020;@Zhang2020a;@Zhang2020c;@VanDerMade2020]. 
Furthermore, aberrant monocyte and neutrophil accumulation along with systemic inflammation 
<span custom-style = 'revision'>and altered systemic iron homeostasis have been linked to the disease severity and a poor prognosis</span> 
[@Schulte-Schrepping2020;@Bellmann-Weiler2021;@Merad2020;@Laing2020;@Lucas2020;@Christensen2021;@Silvin2020;@Penttila2021;@Vitte2020]. This pertains to the key role of 
<span custom-style = 'revision'>myeloid leukocytes, i. e. macrophages, monocytes and neutrophils in the SARS-CoV-2 pathogen control and orchestration of the inflammatory response</span> [@Merad2020;@Mueller2021;@Trombetta2021;@Vasse2021;@Christensen2021;@Bost2021;@Silvin2020].

Human blood monocytes consist of three populations: the major subset of 
<span custom-style = 'revision'>CD14^+/high^ CD16^-/low^ classical cells and substantially</span> 
less abundant CD14^+^ CD16^+^ intermediate and CD14^-/low^ CD16^+/high^ non-classical monocytes [@Wong2012a]. These subsets display differences in the pattern to activating stimuli and 
<span custom-style = 'revision'>immune effector functions</span> 
such as phagocytosis, antigen presentation and cytokine production [@Wong2012a]. Alterations in the monocyte subset 
<span custom-style = 'revision'>distribution and surface molecule repertoire<span> 
were observed in multiple inflammatory conditions including COVID-19 and were proposed to correlate with the disease course and prognosis [@Wong2012a;@Haschka2020;@Patel2017;@Gatti2020;@Penttila2021;@Silvin2020;@Geanon2021;@Schulte-Schrepping2020;@Trombetta2021;@Mueller2021;@Zingaropoli2021;@Christensen2021;@Vasse2021;@Lucas2020].

From the clinical and immunological point of view there is an ongoing search for markers 
<span custom-style = 'revision'>characterizing</span> the clinical course of 
<span custom-style = 'revision'>COVID-19 as well as the protective and pathological immune response to the pathogen.</span>
<span custom-style = 'revision'>Herein we investigated the surface molecule repertoire of myeloid innate immune cells from hospitalized COVID-19 patients by flow cytometry and linked the surface phenotype to the disease severity and patterns of systemic inflammation and iron turnover by self-organizing map clustering[@Kohonen1995; @Vesanto2000].</span>

@@

The study was conducted between March and November 2020 and <span custom-style = 'revision'>encompassed healthy controls (n = 7)</span> and hospitalized COVID-19 patients (n = 53 enrolled) whose blood myeloid compartment was analyzed by flow cytometry within 72 hours from hospital admission <span custom-style = 'revision'>(__Supplementary Figure S1__ and __S2__)</span>. 
<span custom-style = 'revision'>Due to insufficient flow cytometry sample quality, 5 COVID-19 participants were excluded from the analysis. Finally, samples from `r count(suppl_tables$study_dataset, sev_paper)$n[1]` healthy controls, 
`r count(suppl_tables$study_dataset, sev_paper)$n[2]` moderate COVID-19 patients not requiring oxygen therapy during hospitalization (WHO score 3) and `r count(suppl_tables$study_dataset, sev_paper)$n[3]` severe COVID-19 individuals requiring oxygen or mechanical ventilation (WHO score 4 - 6) were available for the analysis (__Figure \@ref(fig:fig-consort)__). Among severe COVID-19 cases, `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), icu)$txt_lab[2]` were treated at an intensive case unit and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), mortality)$txt_lab[2]` deceased during hospitalization.</span> Males constituted `r my_count(filter(cohort$analysis_tbl, sev_paper == 'moderate'), sex)$txt_lab[2]` of the moderate and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), sex)$txt_lab[2]` of the severe COVID-19 group, the sex distribution difference was not significant. 
<span custom-style = 'revision'>The median age was significantly different between the healthy controls (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'healthy'), 'age')` years), moderate (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'moderate'), 'age')` years) and severe COVID-19 patients (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'severe'), 'age')`, years) and peaked in the severe COVID-19 group (__Supplementary Table S1__).</span>

@@

Systemic inflammation, as evidenced by elevated levels of interleukin-6 (IL6), C-reactive protein (CRP) and neopterin, 
<span custom-style = 'revision'>is present in hospitalized COVID-19 individuals and was</span> linked to a poor outcome [@Laing2020; @Lucas2020; @Christensen2021; @Bellmann-Weiler2021]. In our study cohort, <span custom-style = 'revision'>blood IL6, CRP and neopterin concentrations at admission were significantly higher in patients developing severe COVID-19 than in subjects with moderate disease as described previously [@Bellmann-Weiler2021] (__Supplementary Figure S3A__).</span> 
<span custom-style = 'revision'>These differences in inflammatory markers was paralleled by significantly reduced circulating iron concentrations, transferrin saturation and elevated circulating ferritin levels in severe COVID-19 as compared with the moderate group (__Supplementary Figure S3B__). Such phenomenon reflects iron restriction in response to a severe infection, reported for COVID-19 as well [@Weiss2018;@Yagc2021; @Christensen2021; @Brandtner2020; @Bellmann-Weiler2020; @Zhou2020].</span>

@@

Next, we investigated how this systemic inflammatory and <span custom-style = 'revision'>iron-restricted</span> milieu shapes the composition, <span custom-style = 'revision'>inflammatory and iron transporter protein repertoire</span> of myeloid leukocytes [@Recalcati2012; @Nairz2017]. 
<span custom-style = 'revision'>To this end, we analyzed the whole blood monocyte and neutrophil compartment of healthy controls and COVID-19 patients within 72 hours post hospital admission with multi-color flow cytometry (__Supplementary Table S2__). Neutrophils were defined within the lymphocyte lineage-negative leukocyte population (Lin^-^: CD3^-^ CD19^-^ CD56^-^) by logical AND gating of the CD16^+^, CD11b^+^, CD62L^+^ and SSC^hi^ cells. Monocytes were identified within the non-neutrophil gate by UMAP clustering (uniform manifold approximation and projection) [@McInnes2018] in respect to CD11b, CD14, CD15, CD16, CD62L, CCR2, CX3CR1 and HLA-DR expression levels (__Supplementary Figure S1__). Classical, intermediate and non-classical monocyte subsets were defined by differences in CD14, CD16, HLA-DR, CCR2 and CX3CR1 surface levels [@Wong2012a] within the monocyte cluster cells (__Supplementary Figure S2__). Of note, the markers used to define monocyte populations: HLA-DR, CCR2 and CX3CR1 displayed similar expression differences between classical, intermediate and non-classical cells from healthy donors and COVID-19 patients. No downregulation of HLA-DR levels in classical monocytes described previously [@Schulte-Schrepping2020; @Silvin2020; @Christensen2021] could be observed in COVID-19 patients as compared with healthy controls (__Supplementary Figure S4__). In turn, neutrophil CD14 expression was found significantly different between the study groups and was the highest in moderate COVID-19 individuals.</span>

<span custom-style = 'revision'>A total of `r length(cyto_clust$cyto_variables)` features could be measured with our flow cytometry pipeline including the percentages of neutrophils, classical, intermediate and non-classical monocytes as well as isotype staining-controlled surface levels of pro- and anti-inflammatory markers, iron-import protein CD71 (transferrin receptor 1) and iron exporter FPN1 (ferroportin) (__Supplementary Table S3__). Among them, `r length(cyto$kruskal_signif_features)` variables displayed significant differences between healthy controls, moderate and severe COVID-19 cell donors as determined by Kruskal-Wallis test: (1) percentage of lineage-negative cells, (2) neutrophil: lymphocyte ratio (NLR), (3) percentages of non-classical monocytes within the CD45^+^ compartment and (4) pan-monocytes, expression levels of CD64 in (5) neutrophils, (6) classical, (7) intermediate and (8) non-classical monocytes as well as surface CD86 levels in (9) classical and intermediate monocytes (10) (__Figure \@ref(fig:fig-cyto-screen)__).</span>

<span custom-style = 'revision'>More specifically, an expansion of neutrophils at the cost of Lineage-positive lymphocytes could be observed in severe COVID-19 resulting in substantially higher NLR values (median: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'nlr')`) as compared with moderate COVID-19 (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'nlr')`) or healthy controls (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'nlr')`) (__Supplementary Figure S6__). This phenomenon was accompanied by a depletion of the non-classical monocyte subset, which was evident in moderate COVID (median % of pan-monocytes: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_nonclassicalMono')`) and significant in the severe disease (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_nonclassicalMono')`) as compared with healthy cell donors (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_nonclassicalMono')`). In addition, the intermediate monocyte compartment was substantially, yet not significantly larger in COVID-19 patients (median, % of CD45^+^ cells, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_intermediateMono')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'panMono_intermediateMono')`) than in control individuals (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_intermediateMono')`) (__Figure \@ref(fig:fig-subset-distribution)__). Of note, neutrophil expansion [@Silvin2020; @Penttila2021; @Vitte2020; @Laing2020; @Christensen2021] along with a profound reduction of non-classical monocyte levels [@Gatti2020; @Penttila2021; @Silvin2020; @Geanon2021; @Schulte-Schrepping2020; @Trombetta2021; @Mueller2021; @Zingaropoli2021; @Christensen2021; @Vasse2021; @Lucas2020] was described as a characteristic of acute COVID-19.</span>

The neutrophil CD64 was virtually absent in healthy controls (median $\Delta$MFI: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'neutrophils_CD64')`) but highly expressed in moderate (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'neutrophils_CD64')`) and severe COVID-19 (`r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'neutrophils_CD64')`). 
A similar pattern of CD64 regulation was discerned for classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD64')`) and intermediate monocytes (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD64')`), which, however, expressed this FC$\gamma$ receptor at high levels also in non-infected controls <span custom-style = 'revision'>(_Supplementary Figure S7__)</span>.
Surface expression of CD86 in classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD86')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD86')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD86')`) and intermediate monocytes (healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD86')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD86')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD86')`) reached the maximum in moderate COVID-19 patients as compared with healthy controls and severe COVID-19 patients <span custom-style = 'revision'>(__Supplementary Figure S8__)</span>.

@@

<span custom-style = 'revision'>
Healthy controls and moderate COVID-19 cases comprised 75% of the largest participant cluster #1 and was characterized by the lowest levels CD40/CD64/CD71 parameters as well as healthy-like monocyte subset distribution, neutrophil levels and neutrophil CD64 expression (__Figure \@ref(fig:fig-clustering)AB__). Interestingly, none of the severe COVID-19 cases assigned to the cluster #1 required an ICU treatment (__Supplementary Table S4__). 
The remaining clusters consisted predominantly of severe COVID-19 patients. The hallmark of the cluster #2 was an increased expression of CD86, C163, the immunosuppresive PD-L1 molecule (CD279) and the iron exporter FPN1 in classical and intermediate monocytes as well as an expansion of the intermediate subset. The participant cluster #3 in turn was primarily characterized by an elevated neutrophil and monocyte CD64 as well as a depletion of the non-classical monocytes. Finally, leukocyte samples assigned to the cluster #4 displayed a consistent expansion of the neutrophil compartment, elevated CD64 in neutrophils and a depletion non-classical monocytes (__Figure \@ref(fig:fig-clustering)A__).
</span>

<span custom-style = 'revision'>
In a comparison restricted to the COVID-19 individuals, the participant clusters demonstrated similar age and body mass index. In turn, the cluster #1 and #4 COVID-19 individuals were predominantly male (78%) as compared with the remaining subjects. Concerning the disease course, the fraction of participants requiring respiratory support was the lowest in cluster #1 and peaked in the clusters #3 and #4. However, none of these clinical features was significantly different in the all-group and the cluster #1 versus rest comparisons (__Supplementary Table S4__).
</span>

@@

Our study corroborates the massive changes in the neutrophil and monocyte compartment in acute moderate and severe COVID-19 including expansion of neutrophils, depletion of the non-classical monocytes 
<span custom-style = 'revision'>as well as regulation of CD64 and CD86 expression in myeloid leukocytes. Finally, we demonstrate the presence of four patterns of the innate myeloid response to SARS-CoV-2 with possible implication for diagnosis, risk assessment and therapy of COVID-19.</span>.

@@

>>>>

Severe 
<span custom-style = 'revision'>coronavirus disease 19 (COVID-19)</span> 
manifests with systemic immediate pro-inflammatory innate immune activation and altered iron turnover. As iron homeostasis, 
<span custom-style = 'revision'>differentiation and function of myeloid leukocytes</span> 
are interconnected we sought to characterize the cellularity, surface marker expression and iron transporter phenotype of neutrophils and monocyte subsets in COVID-19 patients within 72 hours from hospital admission, and how 
<span custom-style = 'revision'>these parameters</span> 
relate to infection severity.

Between March and November 2020, blood leukocyte samples from 
<span custom-style = 'revision'>hospitalized</span> 
COVID-19 patients 
<span custom-style = 'revision'>(n = 48)</span> 
and healthy individuals 
<span custom-style = 'revision'>(n = 7)</span> 
were analyzed by flow cytometry enabling comparative analysis of 40 features. 
<span custom-style = 'revision'>Inflammation driven neutrophil expansion, depletion of CD16^+^ non-classical monocytes and changes in surface expression of neutrophil and monocyte CD64 and CD86 were associated with COVID-19 severity. By unsupervised self-organizing map clustering, four patterns of innate myeloid response were identified and linked to varying levels of systemic inflammation, altered cellular iron trafficking and the severity of disease.</span> These alterations of the myeloid leukocyte compartment 
<span custom-style = 'revision'>during acute COVID-19</span> 
may be hallmarks of inefficient viral control and  
<span custom-style = 'revision'>immune hyper-activation and may</span> and help at risk prediction and treatment 
<span custom-style = 'revision'>optimization.</span>

@@

Since late 2019, the 
<span custom-style = 'revision'>SARS-CoV-2 pandemic</span> 
has resulted in a global health crisis. 
<span custom-style = 'revision'>As of March 2022, more than 472 million infections and over 6.1 million COVID-19 deaths</span> 
were documented [@WHO2021]. 
COVID-19 is characterized by a heterogeneous clinical manifestation and divergent courses of the disease [@Burkert2021]. This ranges from asymptomatic 
<span custom-style = 'revision'>infection, to ambulatory disease with wide-ranged flu-like complaints, up to severe respiratory or multi-organ failure</span>, 
requiring hospitalization and treatment at intensive care unit [@Sahanic2021; @Oberfeld2020; @Burkert2021]. 
Especially in the latter cases, an excessive over-stimulation of the immune system, likewise driven by an impaired immune control of viral replication leads to a ‘cytokine release syndrome’ with subsequent organ damage [@Riva2020; @Fajgenbaum2020]. Specifically, low levels of CD4^+^ and CD8^+^ cells, 
<span custom-style = 'revision'>impaired anti-viral type I interferon response  
and polymorphisms in the genes involved in antigen presentation and immune cell communication were associated with the risk of severe COVID-19</span> [@Ellinghaus2020; @Zhang2020a; @Zhang2020c; @VanDerMade2020]. 
Furthermore, aberrant monocyte and neutrophil accumulation along with systemic inflammation 
<span custom-style = 'revision'>and altered systemic iron homeostasis have been linked to the disease severity and a poor prognosis</span> 
[@Schulte-Schrepping2020; @Bellmann-Weiler2021; @Merad2020; @Laing2020; @Lucas2020; @Christensen2021; @Silvin2020; @Penttila2021; @Vitte2020; @Bellmann-Weiler2020]. This pertains to the key role of 
<span custom-style = 'revision'>myeloid leukocytes, i. e. macrophages, monocytes and neutrophils in the SARS-CoV-2 pathogen control and orchestration of the inflammatory response</span> [@Merad2020; @Mueller2021; @Trombetta2021; @Vasse2021; @Christensen2021; @Bost2021; @Silvin2020]. 
<span custom-style = 'revision'>In addition, monocytes/macrophages play an important role in the orchestration of iron homeostasis, and inflammatory processes result in macrophage/monocyte iron retention whereas iron exerts distinct effect on macrophage polarization and immune effector functions [@Haschka2021; @Recalcati2012; @Lanser2021].</span>

Human blood monocytes consist of three populations: the major subset of 
<span custom-style = 'revision'>CD14^+/high^ CD16^-/low^ classical cells and substantially</span> 
less abundant CD14^+/high^ CD16^+/high^ intermediate and CD14^-/low^ CD16^+/high^ non-classical monocytes [@Wong2012a]. These subsets display differences in the pattern to activating stimuli and 
<span custom-style = 'revision'>immune effector functions</span> 
such as phagocytosis, antigen presentation and cytokine production [@Wong2012a]. Alterations in the monocyte subset 
<span custom-style = 'revision'>distribution and surface molecule repertoire<span> 
were observed in multiple inflammatory conditions including COVID-19 and were proposed to correlate with the disease course and prognosis [@Wong2012a; @Haschka2020; @Patel2017; @Gatti2020; @Penttila2021; @Silvin2020; @Geanon2021; @Schulte-Schrepping2020; @Trombetta2021; @Mueller2021; @Zingaropoli2021; @Christensen2021; @Vasse2021; @Lucas2020].

From the clinical and immunological point of view there is an ongoing search for markers 
<span custom-style = 'revision'>characterizing</span> the clinical course of 
<span custom-style = 'revision'>COVID-19 as well as the protective and pathological immune response to the pathogen.</span>
<span custom-style = 'revision'>Herein we investigated the surface molecule repertoire of myeloid innate immune cells from hospitalized COVID-19 patients by flow cytometry and linked the surface phenotype to the disease severity and patterns of systemic inflammation and iron turnover by self-organizing map clustering [@Kohonen1995; @Vesanto2000].</span>

@@

The study was conducted between March and November 2020 and <span custom-style = 'revision'>encompassed healthy controls (n = 7)</span> and hospitalized COVID-19 patients (n = 53 enrolled) whose blood myeloid compartment was analyzed by flow cytometry within 72 hours from hospital admission <span custom-style = 'revision'>(__Supplementary Figure S1__ and __S2__)</span>. 
<span custom-style = 'revision'>Due to insufficient flow cytometry sample quality, 5 COVID-19 patients were excluded from further analysis. Finally, samples from `r count(suppl_tables$study_dataset, sev_paper)$n[1]` healthy controls, 
`r count(suppl_tables$study_dataset, sev_paper)$n[2]` moderate COVID-19 patients not requiring supplemental oxygen therapy during hospitalization (WHO score 3) and `r count(suppl_tables$study_dataset, sev_paper)$n[3]` severe COVID-19 individuals requiring low/high flow oxygen treatment or mechanical ventilation (WHO score 4 - 6) were available for the analysis (__Figure \@ref(fig:fig-consort)__). Among severe COVID-19 cases, `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), icu)$txt_lab[2]` were treated at an intensive care unit and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), mortality)$txt_lab[2]` deceased during hospitalization.</span> Males constituted `r my_count(filter(cohort$analysis_tbl, sev_paper == 'moderate'), sex)$txt_lab[2]` of the moderate and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), sex)$txt_lab[2]` of the severe COVID-19 group, the sex distribution difference was not significant. 
<span custom-style = 'revision'>The median age was significantly different between the healthy controls (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'healthy'), 'age')` years), moderate (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'moderate'), 'age')` years) and severe COVID-19 patients (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'severe'), 'age')`, years) and peaked in the severe COVID-19 group (__Supplementary Table S1__).</span>

@@

Systemic inflammation, as evidenced by elevated levels of interleukin-6 (IL6), C-reactive protein (CRP) and neopterin, 
<span custom-style = 'revision'>was present in hospitalized COVID-19 individuals and was</span> linked to a poor outcome [@Laing2020; @Lucas2020; @Christensen2021; @Bellmann-Weiler2021]. In our study cohort, <span custom-style = 'revision'>plasma IL6, CRP and neopterin concentrations at admission were significantly higher in patients developing severe COVID-19 than in subjects with moderate disease as described previously [@Bellmann-Weiler2021] (__Supplementary Figure S3A__).</span> 
<span custom-style = 'revision'>These differences in inflammatory markers was paralleled by significantly reduced circulating iron concentrations, transferrin saturation and elevated circulating ferritin levels in severe COVID-19 as compared with the moderate group (__Supplementary Figure S3B__). Such phenomenon reflects iron restriction in response to a severe infection, reported previously for COVID-19 [@Weiss2018; @Yagc2021; @Christensen2021; @Brandtner2020; @Bellmann-Weiler2020; @Zhou2020].</span>

@@

Next, we investigated how this systemic inflammatory and <span custom-style = 'revision'>iron-restricted</span> milieu shapes the composition, <span custom-style = 'revision'>inflammatory and iron transporter protein repertoire</span> of myeloid leukocytes [@Recalcati2012; @Nairz2017]. 
<span custom-style = 'revision'>To this end, we analyzed the whole blood monocyte and neutrophil compartment of healthy controls and COVID-19 patients within 72 hours of hospital admission with multi-color flow cytometry (__Supplementary Table S2__). Neutrophils were defined within the lymphocyte lineage-negative leukocyte population (Lin^-^: CD3^-^ CD19^-^ CD56^-^) by logical gating (AND) of the CD16^+^, CD11b^+^, CD62L^+^ and SSC^hi^ cells. Monocytes were identified within the non-neutrophil gate by UMAP clustering (uniform manifold approximation and projection) [@McInnes2018] in respect to CD11b, CD14, CD15, CD16, CD62L, CCR2, CX3CR1 and HLA-DR expression levels (__Supplementary Figure S1__). Classical, intermediate and non-classical monocyte subsets were defined by differences in CD14, CD16, HLA-DR, CCR2 and CX3CR1 surface levels [@Wong2012a] within the monocyte cluster cells (__Supplementary Figure S2__). Of note, the additional markers used to define monocyte populations (HLA-DR, CCR2 and CX3CR1) displayed similar expression differences between classical, intermediate and non-classical cells from healthy donors and COVID-19 patients. No downregulation of HLA-DR levels in classical monocytes, described previously [@Schulte-Schrepping2020; @Silvin2020; @Christensen2021], could be observed in COVID-19 patients as compared with healthy controls (__Supplementary Figure S4__). In turn, neutrophil CD14 expression was found significantly different between the study groups and was the highest in moderate COVID-19 individuals (__Supplementary Figure S5__).</span>

<span custom-style = 'revision'>A total of `r length(cyto_clust$cyto_variables)` features could be measured with our flow cytometry pipeline including the percentages of neutrophils, classical, intermediate and non-classical monocytes as well as isotype staining-controlled surface levels of pro- and anti-inflammatory markers, iron-import protein CD71 (transferrin receptor 1) and iron exporter FPN1 (ferroportin) (__Supplementary Table S3__). Among them, `r length(cyto$kruskal_signif_features)` variables displayed significant differences between healthy controls, moderate and severe COVID-19 cell donors as determined by Kruskal-Wallis test: (1) percentage of lineage-negative cells, (2) neutrophil: lymphocyte ratio (NLR), (3) percentages of non-classical monocytes within the CD45^+^ compartment and (4) pan-monocytes, expression levels of CD64 in (5) neutrophils, (6) classical, (7) intermediate and (8) non-classical monocytes as well as surface CD86 levels in (9) classical and (10) intermediate monocytes (__Figure \@ref(fig:fig-cyto-screen)__).</span>

<span custom-style = 'revision'>More specifically, an expansion of neutrophils at the cost of Lineage-positive lymphocytes could be observed in severe COVID-19 resulting in substantially higher NLR values (median: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'nlr')`) as compared with moderate COVID-19 (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'nlr')`) or healthy controls (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'nlr')`) (__Supplementary Figure S6__). This phenomenon was accompanied by a depletion of the non-classical monocyte subset, which was evident in moderate COVID (median % of pan-monocytes: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_nonclassicalMono')`) and significant in the severe disease (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_nonclassicalMono')`) as compared with healthy cell donors (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_nonclassicalMono')`). In addition, the intermediate monocyte compartment was substantially, yet not significantly, larger in COVID-19 patients (median, % of CD45^+^ cells, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_intermediateMono')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'panMono_intermediateMono')`) than in control individuals (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_intermediateMono')`) (__Figure \@ref(fig:fig-subset-distribution)__). Of note, neutrophil expansion [@Silvin2020; @Penttila2021; @Vitte2020; @Laing2020; @Christensen2021] along with a profound reduction of non-classical monocyte levels [@Gatti2020; @Penttila2021; @Silvin2020; @Geanon2021; @Schulte-Schrepping2020; @Trombetta2021; @Mueller2021; @Zingaropoli2021; @Christensen2021; @Vasse2021; @Lucas2020] has been described as a characteristic of acute COVID-19.</span>

The neutrophil CD64 was virtually absent in healthy controls (median $\Delta$MFI: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'neutrophils_CD64')`) but highly expressed in moderate (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'neutrophils_CD64')`) and severe COVID-19 (`r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'neutrophils_CD64')`). 
A similar pattern of CD64 regulation was discerned for classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD64')`) and intermediate monocytes (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD64')`), which, however, expressed this FC$\gamma$ receptor at high levels also in non-infected controls <span custom-style = 'revision'>(__Supplementary Figure S7__)</span>.
Surface expression of CD86 in classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD86')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD86')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD86')`) and intermediate monocytes (healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD86')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD86')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD86')`) reached the maximum in moderate COVID-19 patients as compared with healthy controls and severe COVID-19 patients <span custom-style = 'revision'>(__Supplementary Figure S8__)</span>.

@@

<span custom-style = 'revision'>
Healthy controls and moderate COVID-19 cases comprised 75% of the largest participant cluster #1 that was characterized by the lowest levels CD40/CD64/CD71 parameters as well as healthy-like monocyte subset distribution, neutrophil levels and neutrophil CD64 expression (__Figure \@ref(fig:fig-clustering)AB__). Interestingly, none of the severe COVID-19 cases assigned to the cluster #1 required an ICU treatment (__Supplementary Table S4__). 
The remaining clusters consisted predominantly of severe COVID-19 patients with increasing frequency. The hallmark of the cluster #2 was an increased expression of CD86, C163, the immunosuppressive PD-L1 molecule (CD279) and the iron exporter FPN1 in classical and intermediate monocytes as well as an expansion of the intermediate subset. The participant cluster #3 in turn was primarily characterized by an elevated neutrophil and monocyte CD64 as well as a depletion of the non-classical monocytes. Finally, leukocyte samples assigned to the cluster #4 displayed a consistent expansion of the neutrophil compartment, elevated CD64 in neutrophils and a depletion non-classical monocytes (__Figure \@ref(fig:fig-clustering)A__).
</span>

<span custom-style = 'revision'>
In comparative analysis of COVID-19 individuals, the participant clusters demonstrated similar age and body mass index. In turn, the cluster #1 and #4 COVID-19 individuals were predominantly male (78%) as compared with the remaining subjects. Concerning the disease course, the fraction of participants requiring respiratory support was the lowest in cluster #1 and peaked in the clusters #3 and #4. However, none of these clinical features was significantly different in the all-group and the cluster #1 versus rest comparisons (__Supplementary Table S4__).
</span>

@@

Our study corroborates the massive changes in the neutrophil and monocyte compartment in acute moderate and severe COVID-19 including expansion of neutrophils, depletion of the non-classical monocytes 
<span custom-style = 'revision'>as well as regulation of CD64 and CD86 expression and altered iron trafficking in myeloid leukocytes. Finally, we demonstrate the presence of four patterns of the innate myeloid response to SARS-CoV-2 with possible implication for diagnosis, risk assessment and therapy of COVID-19.</span>

@@

## Measurement of laboratory inflammation and iron markers
